% !TeX program = pdflatex
% The Pentagon, Machine-Checked — Abel's five-term relation for the dilogarithm in Lean 4.
% Sequel to "The Clock That Never Ticks": closes the five-term thread that paper left open.
% Build: pdflatex five-term-lean.tex (twice, for refs)
% Verified against the Lean source 2026-06-13 (lake build Dilog.FiveTerm, exit 0;
% #print axioms on every named theorem -> propext, Classical.choice, Quot.sound only).
% Toolchain v4.30.0-rc2, Mathlib pin 701fb6e9.
% All numerical instances re-verified to 30 dps (research/_tmp).

\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\DeclareUnicodeCharacter{2082}{\ensuremath{_2}}
\usepackage{amsmath,amssymb,amsthm,booktabs,graphicx,hyperref}
\usepackage{xurl}
\usepackage[table]{xcolor}
\definecolor{ggteal}{HTML}{1B6F8C}
\definecolor{ggband}{HTML}{DCE7EB}
\definecolor{ggamber}{HTML}{E08A1E}
\usepackage[margin=1in]{geometry}
\usepackage{microtype}
\usepackage{float}
\usepackage[section]{placeins}
\usepackage[most]{tcolorbox}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning}
\setlength{\emergencystretch}{2em}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{remark}{Remark}
\newcommand{\lean}[1]{\texttt{#1}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Li}{\mathrm{Li}_2}
\DeclareMathOperator{\Lrog}{L}
\hypersetup{colorlinks=true,linkcolor=ggteal,citecolor=ggteal,urlcolor=ggteal}
\newtcbox{\keybox}{on line,boxrule=0.9pt,colframe=ggteal,colback=ggband!40,
  arc=3pt,left=8pt,right=8pt,top=5pt,bottom=5pt}
\newcommand{\keyeq}[1]{\begin{center}\keybox{$\displaystyle #1$}\end{center}}
\newtcolorbox{recipebox}[1]{colframe=ggamber,colback=ggamber!8,arc=3pt,boxrule=0.8pt,
  title=\textbf{#1},coltitle=white,colbacktitle=ggamber}

\title{\bf The Pentagon, Machine-Checked\\[4pt]
  \large\normalfont Abel's five-term relation for the dilogarithm in Lean~4}
\author{Carles Mar\'in\thanks{Formalized with AI assistance (Claude, Anthropic); the
mathematics and all claims are the author's responsibility. The methodology and the exact
boundary of the contribution are in Section~\ref{sec:boundary}.}\\[3pt]
  \normalsize Independent researcher\\
  \normalsize \texttt{karlesmarin@gmail.com}}
\date{June 2026}

\begin{document}
\maketitle

\begin{abstract}
I give a machine-checked, \lean{sorry}-free proof, in Lean~4 over Mathlib, of \emph{Abel's
five-term relation} for the dilogarithm. With $\Lrog$ the Rogers $\Lrog$-function it reads,
for $x,y\in(0,1)$,
\[
  \Lrog(x)+\Lrog(y)=\Lrog(xy)+\Lrog\!\Big(\tfrac{x(1-y)}{1-xy}\Big)
    +\Lrog\!\Big(\tfrac{y(1-x)}{1-xy}\Big).
\]
This single identity is the whole weight-2 theory in one line: by theorems of Wojtkowiak
and de~Jeu, \emph{every} functional equation of $\Li$ with rational arguments is a
consequence of it; it is the defining relation of the Bloch group and the pentagon of
cluster algebras and of the quantum dilogarithm. A companion paper, \emph{The Clock That
Never Ticks}, built a dilogarithm in Lean from scratch and climbed the golden-ratio ladder
\emph{without} this relation, leaving it explicitly open; this paper closes that thread.
The proof is the classical differentiation argument made fully rigorous: fixing $y$, the
five-term difference has identically zero derivative on $(0,1)$---nine logarithmic terms
collapse into a basis of five and the rest is \lean{ring}---so it is constant, and its
boundary value as $x\to0^+$ is computed by squeezing $\log x\cdot\log(1-x)\to0$. I am exact
about the boundary: the mathematics is entirely classical (Spence, Abel, Rogers;
Wojtkowiak's and de~Jeu's universality); the contribution is the formalization---to the
best of my knowledge the first machine-checked five-term relation, on the first
machine-checked dilogarithm, in any proof assistant---together with three reusable building
blocks. The headline theorem and its blocks depend only on \lean{propext},
\lean{Classical.choice} and \lean{Quot.sound}.
\end{abstract}

\medskip
\noindent\emph{How to read this paper.} It is the story of four questions, each opening a
section and handing the reader the next, walking from a puzzle to a closed thread. The
puzzle: a companion paper reached the dilogarithm's deepest evaluations \emph{without} the
identity everyone calls central---so is the identity worth machine-checking at all
(Q1)? It is, because it is not one more identity but the \emph{generator} of all of them;
that earns it the proof (Q2), a proof the machine sharpened in one place a pen would skate
over (Q3); and once it is certified, the showpieces the companion reached the long way round
fall out as corollaries (Q4). If you read only one thing, read Theorem~\ref{thm:five} and
Figure~\ref{fig:pentagon}; the Lean inventory is Table~\ref{tab:inventory}; the honest
limits are in Section~\ref{sec:boundary}; reproduction is three commands.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  obj/.style={draw=ggteal,thick,rounded corners=2pt,fill=ggband!40,align=center,
    font=\small,inner sep=5pt,minimum height=10mm},
  q/.style={font=\scriptsize\itshape,text=ggamber!80!black,align=center},
  arr/.style={-{Stealth[length=2.4mm]},thick,ggteal}]
  \node[obj] (clock) {the Clock paper\\climbs without Abel};
  \node[obj,right=24mm of clock] (gen) {the five-term is\\the \emph{generator}};
  \node[obj,right=24mm of gen] (proof) {three moves:\\$G'\!=\!0$, const, boundary};
  \node[obj,below=15mm of proof] (caught) {what the machine\\caught};
  \node[obj,left=24mm of caught] (coro) {ladders \emph{with} Abel,\\Bloch, cluster};
  \node[obj,left=24mm of coro,fill=ggamber!25,draw=ggamber] (done)
    {\textbf{the pentagon,}\\\textbf{machine-checked}};
  \draw[arr] (clock) -- node[q,above=1mm] {Q1: worth\\checking?} (gen);
  \draw[arr] (gen) -- node[q,above=1mm] {Q2: same humble\\recipe?} (proof);
  \draw[arr] (proof) -- node[q,right=1mm] {Q3: what did\\the pen skate?} (caught);
  \draw[arr] (caught) -- node[q,above=1mm] {Q4: what falls\\out free?} (coro);
  \draw[arr] (coro) -- (done);
\end{tikzpicture}
\caption{The reader's map. The thread starts where the companion \emph{Clock} paper left it
(the golden ladder climbed without Abel) and closes at the machine-checked pentagon; each
arrow is the question its section answers. Every node is a block of \lean{sorry}-free Lean.}
\label{fig:readermap}
\end{figure}

\section{Q1: why machine-check an identity the companion paper did without?}
\label{sec:intro}

Some identities matter for what they assert; this one matters for where it keeps
reappearing. Take the five numbers that sit at the corners of a pentagon, apply the
dilogarithm to each, and the five values are not free: they obey one fixed relation, the
same one whatever pentagon you start from. That single fact is, in disguise, the defining
relation of the Bloch group in hyperbolic geometry, the pentagon of mutations in cluster
algebra, and the pentagon identity of the quantum dilogarithm in mathematical
physics---one relation wearing three masks. It is two centuries old and has been
rediscovered under a dozen names; and yet, after a decade of one of the most ambitious
formalization efforts ever attempted, no proof assistant had ever been told it is true. This
paper tells one---and then asks why such a thing should be everywhere at once.

A taste of what the relation does, before any theory. Put $x=y=\tfrac12$. After
$\tfrac{(1/2)(1/2)}{1-1/4}=\tfrac13$ the relation below collapses to
\[
  2\,\Lrog\!\big(\tfrac12\big)=\Lrog\!\big(\tfrac14\big)+2\,\Lrog\!\big(\tfrac13\big),
  \qquad\text{i.e.}\qquad \Lrog\!\big(\tfrac14\big)+2\,\Lrog\!\big(\tfrac13\big)=\frac{\pi^2}{6},
\]
since $\Lrog(\tfrac12)=\pi^2/12$. That a sum of three dilogarithm values at $\tfrac14$ and
$\tfrac13$ is \emph{exactly} $\pi^2/6$ is not obvious; it is one of infinitely many shadows
of a single two-variable law---and that law is what we machine-check here.

\emph{History bead.} The dilogarithm is Euler's (1760s). Its functional equations were
charted by Landen (1780), Spence (1809), Abel (1827, published 1881~\cite{abel}) and
Kummer; Rogers gave the symmetric form in 1907~\cite{rogers}. The five-term relation's depth
became structural only later, with Bloch's regulators~\cite{bloch} and the group that bears
his name.

There is a reason to hesitate. A companion paper in this line, \emph{The Clock That Never
Ticks}~\cite{clock}, built a dilogarithm in Lean from scratch---its series, its derivative
$-\log(1-x)/x$, Euler's reflection, Landen's transformation, the duplication
formula---and then reached the subject's classical showpieces, the golden-ratio evaluations
$\Li(1/\varphi)$ and $\Li(1/\varphi^2)$, by a three-by-three linear system, with \emph{no
five-term relation at all}. Every textbook reaches those values through Abel; the
formalization found a side door. So the honest first question is not how to prove the
five-term relation but whether to bother: if the showpieces are reachable without it, what
does the identity earn its keep on?

The answer is that the five-term relation is not one identity among several. It is the
\emph{generator} of the whole subject. Wojtkowiak proved that every one-variable functional
equation of $\Li$ with rational arguments is a $\mathbb{Z}$-linear combination of instances
of the five-term relation; de~Jeu extended this to any number of
variables~\cite{wojtkowiak,dejeu}. The same relation \emph{defines} the Bloch group, hence
governs scissors-congruence invariants and the volumes of hyperbolic
$3$-manifolds~\cite{bloch,suslin,neumannzagier}; in cluster algebras it is the
\emph{pentagon}, the relation of five mutations~\cite{nakanishi}; and in its quantum form it
is the pentagon identity of the Faddeev--Kashaev quantum dilogarithm~\cite{faddeevkashaev}.
The companion's side door climbs one ladder; the five-term relation is the staircase. That
is what is worth certifying. \emph{Can the same humble toolkit that proved reflection and
Landen also reach the generator?}

The cleanest statement uses the \emph{Rogers $\Lrog$-function}
$\Lrog(x):=\Li(x)+\tfrac12\log x\log(1-x)$, which absorbs the stray logarithms of the raw
$\Li$ form and turns the relation into a pure sum of five $\Lrog$ values:
\keyeq{\Lrog(x)+\Lrog(y)=\Lrog(xy)+\Lrog\!\Big(\tfrac{x(1-y)}{1-xy}\Big)
  +\Lrog\!\Big(\tfrac{y(1-x)}{1-xy}\Big),\qquad x,y\in(0,1).}

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=1.5,every node/.style={font=\small}]
  \foreach \i/\lab in {0/{$x$},1/{$\tfrac{x(1-y)}{1-xy}$},2/{$y$},
      3/{$\tfrac{y(1-x)}{1-xy}$},4/{$xy$}}{
    \node (n\i) at ({90+\i*72}:1.25) {};
    \node at ({90+\i*72}:1.62) {\lab};
    \fill[ggteal] ({90+\i*72}:1.25) circle (1.5pt);
  }
  \foreach \i in {0,...,4}{
    \pgfmathtruncatemacro{\j}{mod(\i+1,5)}
    \draw[ggteal,-{Stealth[length=4pt]}] (n\i) -- (n\j);
  }
\end{tikzpicture}
\caption{\label{fig:pentagon}The five arguments on a cycle. The five-fold symmetry of this
pentagon is the combinatorial shadow shared by the Bloch-group relation, the
cluster-algebra mutation pentagon, and the quantum-dilogarithm pentagon identity.}
\end{figure}

\section{Q2: can the generator be reached by the same three-move recipe?}\label{sec:proof}

\emph{History bead.} The differentiation proof is the oldest one: fix one variable, show
the derivative in the other vanishes, evaluate the constant at an endpoint. It is
essentially Abel's, repeated in Lewin~\cite{lewin} and Zagier~\cite{zagier}. The whole
question is whether ``the derivative vanishes'' and ``evaluate at the endpoint'' survive
being made fully formal.

\begin{theorem}[\lean{Dilog.rogersL\_fiveterm}]\label{thm:five}
For all $x,y\in(0,1)$,
\[
  \Lrog(x)+\Lrog(y)=\Lrog(xy)+\Lrog\!\Big(\tfrac{x(1-y)}{1-xy}\Big)
    +\Lrog\!\Big(\tfrac{y(1-x)}{1-xy}\Big).
\]
\end{theorem}

The recipe is the one the companion paper used for every elementary identity: differentiate,
conclude constant, evaluate at the boundary. Fix $y\in(0,1)$ and set
\[
  G(t)=\Lrog(t)-\Lrog(ty)-\Lrog\!\Big(\tfrac{t(1-y)}{1-ty}\Big)
    -\Lrog\!\Big(\tfrac{y(1-t)}{1-ty}\Big).
\]
The theorem is $G(x)=-\Lrog(y)$. Three movements.

\paragraph{1. The derivative vanishes.} For $x,y\in(0,1)$ all three composite arguments
again lie in $(0,1)$, so every term is differentiable. With $\Lrog'(t)=-\tfrac12\big(
\tfrac{\log(1-t)}{t}+\tfrac{\log t}{1-t}\big)$, the chain rule writes $G'(t)$ as four such
expressions weighted by the derivatives of the composite arguments. Every logarithm that
occurs---of $t$, of $ty$, of the two fractional arguments and their complements---expands,
by $\log(ab)=\log a+\log b$ and $\log(a/b)=\log a-\log b$, into the five-element basis
$\{\log t,\log y,\log(1-t),\log(1-y),\log(1-ty)\}$. In that basis $G'(t)$ is a rational
expression whose every coefficient cancels: $G'\equiv0$. In Lean this is the one delicate
step (\lean{fiveterm\_hasDerivAt\_zero}): the nine logarithms are rewritten into the basis
and the rational remainder is cleared and closed by \lean{ring}.

\paragraph{2. Hence $G$ is constant.} A function with vanishing derivative on the open
connected interval $(0,1)$ is constant there---Mathlib's
\lean{IsOpen.is\_const\_of\_deriv\_eq\_zero}, the mean-value theorem in disguise. So
$G(x)=G(t)$ for every $t\in(0,1)$.

\paragraph{3. The boundary value.} As $t\to0^+$, $ty\to0$, $\tfrac{t(1-y)}{1-ty}\to0$ and
$\tfrac{y(1-t)}{1-ty}\to y$, so $G(t)\to\Lrog(0)-\Lrog(0)-\Lrog(0)-\Lrog(y)=-\Lrog(y)$---%
\emph{provided} $\Lrog$ is right-continuous at $0$ with $\Lrog(0)=0$. This is the one
analytic subtlety, the same one the companion paper isolated. $\Lrog(t)=\Li(t)+\tfrac12\log
t\log(1-t)$, and while $\Li$ is continuous, $\log t\cdot\log(1-t)$ is an indeterminate
$(-\infty)\cdot 0$. It is tamed by a squeeze: on $(0,\tfrac12]$ one has $-\log(1-t)\le 2t$
(from $\log z\ge1-z^{-1}$), so
\[
  0\ \le\ \log t\cdot\log(1-t)=(-\log t)\,(-\log(1-t))\ \le\ 2t\,(-\log t)=2\,(-t\log t)
  \ \longrightarrow\ 0,
\]
because $t\log t\to0$ (Mathlib's \lean{Real.continuous\_mul\_log}). Thus $G\to-\Lrog(y)$;
being constant and equal to $G(x)$ near $0$, it \emph{equals} $-\Lrog(y)$, which is
Theorem~\ref{thm:five}. \qed

So the answer to Q2 is yes: the generator needs no machinery the elementary identities did
not---only that the nine-into-five cancellation actually close. \emph{And there the pen and
the machine parted company.}

\section{Q3: what did the machine catch that the pen skates over?}\label{sec:caught}

\emph{History bead.} ``$G'\equiv0$ by a short calculation'' is how every textbook dispatches
step~1. The phrase hides a clearing of denominators that, on paper, one trusts. A proof
assistant does not trust it; it executes it, and execution has an order.

The cancellation was checked numerically to thirty digits before it was formalized---and the
formalization still caught a subtlety the eye misses. When \lean{field\_simp} clears the
denominators of $G'(t)$ it silently reorders the product $x\cdot y$ to $y\cdot x$, and then
needs the hypothesis $1-y x\neq0$---in that exact term order---present in context to
discharge the inverse it produces. Supplied a moment too late, the goal will not close, and
no amount of \lean{ring} repairs it; supplied in the wrong order ($1-xy$ rather than
$1-yx$), likewise. The fix is one line, but finding it cost more than the rest of the proof:
the non-vanishing fact had to be stated for both orders and placed before the
simplification, not after.

\begin{recipebox}{What a proof assistant earns its keep on}
On paper, ``clear denominators and the rest cancels'' is one phrase. Formally it is an
ordered rewrite: \lean{field\_simp} may reorder a product ($x y \rightsquigarrow y x$) and
then demand the matching non-vanishing fact ($1-yx\neq0$) \emph{in that order, already in
context}. The diagnostic, banked for the next weight-2 development: if \lean{field\_simp}
leaves an uncleared $(1-a)^{-1}$, the missing hypothesis is $(1-a)\neq0$ in the term order
\lean{field\_simp} produced---supply both $1-xy\neq0$ and its commute $1-yx\neq0$.
\end{recipebox}

This is the whole of the new mathematics, and it is honestly small: not a theorem but a
discipline. It is also exactly the kind of step that justifies machine-checking an identity
whose informal proof everyone already believes. \emph{With the generator certified, what
does the companion paper's long way round now buy us for free?}

\section{Q4: with the pentagon certified, what falls out for free?}\label{sec:corollaries}

\emph{History bead.} The golden-ratio evaluations are Landen's (1780) and Rogers'
(1907~\cite{rogers}); their modern role as conformal central charges belongs to the
thermodynamic Bethe ansatz of the 1990s~\cite{zamolodchikov,nahm,kirillov}. Every classical
route to them passes through the five-term relation; the companion paper took a detour, and
now the main road is open.

The five-term relation is a machine that turns a choice of $(x,y)$ into a closed relation
among five $\Lrog$ values. Feed it the right points and the catalogue of the subject prints
out (Table~\ref{tab:instances}). Two instances are especially clean because their arguments
\emph{coincide}. At $(x,y)=(\tfrac12,\tfrac12)$ the two fractional arguments are both
$\tfrac13$, giving $2\Lrog(\tfrac12)=\Lrog(\tfrac14)+2\Lrog(\tfrac13)$. At
$(x,y)=(1/\varphi,1/\varphi)$, the three composite arguments \emph{all} collapse to
$1/\varphi^2$ (because $1-1/\varphi^2=1/\varphi$ and $1-1/\varphi=1/\varphi^2$), so the
five-term relation reads $2\Lrog(1/\varphi)=3\Lrog(1/\varphi^2)$---and with the ladder
values $\Lrog(1/\varphi)=\pi^2/10$, $\Lrog(1/\varphi^2)=\pi^2/15$ this is the exact identity
$\pi^2/5=\pi^2/5$. The golden ladder the companion reached by a side door is, from here, a
single substitution.

\begin{table}[h]
\centering\small
\rowcolors{2}{ggband!35}{white}
\begin{tabular}{@{}llll@{}}
\toprule
\rowcolor{ggteal!18} $(x,y)$ & arguments & the relation reads & exact value \\
\midrule
$(\tfrac12,\tfrac12)$ & $\tfrac14,\tfrac13,\tfrac13$
  & $2\Lrog(\tfrac12)=\Lrog(\tfrac14)+2\Lrog(\tfrac13)$
  & $\Lrog(\tfrac14)+2\Lrog(\tfrac13)=\tfrac{\pi^2}{6}$ \\
$(\tfrac1\varphi,\tfrac1\varphi)$ & $\tfrac1{\varphi^2},\tfrac1{\varphi^2},\tfrac1{\varphi^2}$
  & $2\Lrog(\tfrac1\varphi)=3\Lrog(\tfrac1{\varphi^2})$
  & $\tfrac{\pi^2}{5}=\tfrac{\pi^2}{5}$ \\
\bottomrule
\end{tabular}
\caption{The generator at work: two choices of $(x,y)$ whose composite arguments coincide,
and the exact identity each prints. Both rows are re-verified to thirty digits; the first is
hand-checkable from $\Lrog(\tfrac12)=\pi^2/12$, the second from the ladder values
$\Lrog(1/\varphi)=\pi^2/10$, $\Lrog(1/\varphi^2)=\pi^2/15$. Euler's reflection
$\Lrog(x)+\Lrog(1-x)=\pi^2/6$ is \emph{not} an instance of the five-term relation but its
companion---the symmetrization that makes the Rogers form clean---and is proved separately.}
\label{tab:instances}
\end{table}

\begin{recipebox}{The recipe: read off a $\pi^2$-evaluation in three lines}
Pick any $x,y\in(0,1)$ at which the four arguments are nice. (1) Form
$xy,\ \tfrac{x(1-y)}{1-xy},\ \tfrac{y(1-x)}{1-xy}$. (2) Apply
$\Lrog(x)+\Lrog(y)=$ their sum, with a machine-checked guarantee. (3) Substitute any
$\Lrog$ values you know.\\[3pt]
\emph{Worked:} take $x=y=\tfrac12$. Then $xy=\tfrac14$ and both fractions are $\tfrac13$,
so $2\,\Lrog(\tfrac12)=\Lrog(\tfrac14)+2\,\Lrog(\tfrac13)$. Since $\Lrog(\tfrac12)=
\Li(\tfrac12)+\tfrac12\log^2\!2$ and $\Li(\tfrac12)=\tfrac{\pi^2}{12}-\tfrac12\log^2\!2$,
the log terms cancel and $\Lrog(\tfrac12)=\tfrac{\pi^2}{12}$; hence
$\Lrog(\tfrac14)+2\,\Lrog(\tfrac13)=\tfrac{\pi^2}{6}$---a closed form for three dilogarithm
values, certified.
\end{recipebox}

So the thread the companion paper left hanging is closed. It climbed the golden ladder
without Abel and asked whether the side door was a curiosity or a necessity; the answer is
that it was a shortcut, and the staircase it bypassed is now machine-checked. The four
questions end where they began---at a sum of dilogarithm values equal to $\pi^2/6$---but the
identity behind it is no longer a shadow on the wall: it is a theorem in the library, with
the whole weight-2 subject hanging from it.

\section{The building blocks}\label{sec:blocks}

The headline rests on three reusable pieces, each \lean{sorry}-free with the standard three
axioms (Table~\ref{tab:inventory}).

\begin{lemma}[\lean{Dilog.hasDerivAt\_rogersL}]
On $(0,1)$, $\displaystyle\Lrog'(x)=-\tfrac12\Big(\tfrac{\log(1-x)}{x}+\tfrac{\log
x}{1-x}\Big)$.
\end{lemma}
The same derivative scheme the companion used for reflection and Landen: combine
$\Li'(x)=-\log(1-x)/x$ with the product rule for $\tfrac12\log x\log(1-x)$.

\begin{lemma}[\lean{Dilog.tendsto\_rogersL\_nhdsGT\_zero}]
$\Lrog(t)\to0$ as $t\to0^+$.
\end{lemma}
The $\Li$ part is continuous on $[-1,1]$; the log-product is squeezed to $0$ by
$2(-t\log t)$. This is the only place the proof leaves pure algebra for analysis.

\begin{lemma}[\lean{Dilog.continuousAt\_rogersL}]
$\Lrog$ is continuous at every $x\in(0,1)$.
\end{lemma}
Needed for the one surviving boundary term, $\Lrog\big(\tfrac{y(1-t)}{1-ty}\big)\to
\Lrog(y)$.

\begin{table}[h]
\centering\small
\rowcolors{2}{ggband!35}{white}
\begin{tabular}{@{}lll@{}}
\toprule
\rowcolor{ggteal!18} Lean name & statement & where \\
\midrule
\lean{rogersL\_fiveterm} & the five-term relation (Theorem~\ref{thm:five}) & \S\ref{sec:proof} \\
\lean{fiveterm\_hasDerivAt\_zero} & $G'\equiv0$: nine logs collapse to five & \S\ref{sec:proof}--\ref{sec:caught} \\
\lean{hasDerivAt\_rogersL} & $\Lrog'(x)=-\tfrac12(\tfrac{\log(1-x)}{x}+\tfrac{\log x}{1-x})$ & \S\ref{sec:blocks} \\
\lean{tendsto\_rogersL\_nhdsGT\_zero} & $\Lrog(t)\to0$ as $t\to0^+$ & \S\ref{sec:blocks} \\
\lean{continuousAt\_rogersL} & $\Lrog$ continuous on $(0,1)$ & \S\ref{sec:blocks} \\
\bottomrule
\end{tabular}
\caption{The load-bearing results, all \lean{sorry}-free in \lean{Dilog/FiveTerm.lean} of
the repository, on the dilogarithm of \lean{Dilog/Basic.lean} from the companion paper.
Axioms of every entry: \lean{propext}, \lean{Classical.choice}, \lean{Quot.sound} only.}
\label{tab:inventory}
\end{table}

\section{The boundary of the contribution}\label{sec:boundary}

Everything here that is mathematics is classical. The five-term relation is Spence's and
Abel's; the Rogers $\Lrog$-function is Rogers's (1907); the universality statements are
Wojtkowiak's and de~Jeu's; the differentiation proof is standard. I claim no new identity
and no new inequality. The contributions are three: the \emph{formalization} (to my
knowledge the first machine-checked five-term relation, on the first machine-checked
dilogarithm, in any proof assistant); the \emph{reusable infrastructure}---the
Rogers-$\Lrog$ derivative, the right-continuity at $0$ via the $-t\log t$ squeeze, and the
interior continuity---which is exactly what any further weight-2 development (ladders, the
Bloch group, the Bloch--Wigner five-term) will need; and the \emph{proof-engineering record}
of Section~\ref{sec:caught}, where the nine-into-five cancellation is the kind of step a
formal check earns its keep on.

\emph{On methodology.} The formalization was done with an AI assistant (Claude, Anthropic),
under the discipline of the companion paper: the central cancellation verified numerically
to thirty digits before formalizing, and every named theorem checked by \lean{\#print
axioms} to depend only on \lean{propext}, \lean{Classical.choice} and \lean{Quot.sound}.
Correctness does not rest on the assistant: the Lean kernel checks every proof. The
mathematics and the claims are the author's responsibility.

\section{Open questions}\label{sec:questions}

The pentagon is now a theorem in the library. Standing on it, six questions worth the next
effort---the last one is the one I would keep.

\begin{enumerate}
\item \emph{Ladders as a generated family.} Each row of Table~\ref{tab:instances} is the
five-term relation evaluated at a point; mechanizing the substitution so that every ladder
\emph{prints} from the one theorem would turn a catalogue into a corollary. How much of the
classical table of dilogarithm values is a finite orbit of the five-term relation under
$\mathbb{Q}$-rational substitution---and where does that orbit stop?
\item \emph{The Bloch--Wigner sequel.} The same relation holds for the single-valued
Bloch--Wigner function on $\mathbb{C}$, where it underlies hyperbolic volume and scissors
congruence~\cite{suslin,neumannzagier,goncharov}. What is the smallest faithful Lean account
of the Bloch group in which $\mathrm{Vol}$ of a hyperbolic $3$-manifold becomes a computable,
certified number?
\item \emph{The quantum limit.} The Faddeev--Kashaev quantum dilogarithm satisfies a
pentagon identity~\cite{faddeevkashaev} whose $\hbar\to0$ limit is the relation proved here.
Can a proof assistant make that limit a theorem rather than a heuristic---and would doing so
expose, as the classical proof did, a step the physics literature skates over?
\item \emph{Universality, formalized.} Wojtkowiak and de~Jeu reduce \emph{every} rational
functional equation of $\Li$ to the five-term relation~\cite{wojtkowiak,dejeu}. That
reduction is an algorithm. What would it take to run it inside Lean, so that an arbitrary
claimed dilogarithm identity is \emph{decided} against the generator?
\item \emph{What the machine sees and the page hides.} Section~\ref{sec:caught} found one
place where the formal proof was strictly sharper than the informal one---an ordered rewrite
the pen trusts. In a subject two centuries old and re-proved a dozen times, how many such
silent steps remain, and is ``the proof a human believes'' ever quite the proof a machine
will accept?
\item \emph{One to keep.} The five-term relation says five dilogarithm values, read around a
pentagon, sum to nothing new. It is the defining relation of the Bloch group, of cluster
mutation, of the quantum dilogarithm---one identity wearing the masks of geometry, algebra,
and physics. Now that it is machine-checked, the question is not whether it is true but why
it is \emph{everywhere}: what is it about a pentagon that so many parts of mathematics are,
underneath, the same five steps that return you home?
\end{enumerate}

\section*{Reproducing}
\begin{verbatim}
git clone https://github.com/karlesmarin/godsil-gutman-lean
cd godsil-gutman-lean && lake exe cache get
lake build Dilog.FiveTerm
\end{verbatim}
Lean \lean{v4.30.0-rc2}, Mathlib pin \lean{701fb6e9}. The relation is
\lean{Dilog.rogersL\_fiveterm} in \lean{Dilog/FiveTerm.lean}; its supports are in
Table~\ref{tab:inventory}. Each is checked by \lean{\#print axioms} to depend only on
\lean{propext}, \lean{Classical.choice}, \lean{Quot.sound}.

\begin{thebibliography}{99}\small
\bibitem{abel} N.~H.~Abel, \emph{\OE uvres compl\`etes} (L.~Sylow, S.~Lie, eds.), vol.~II,
Gr{\o}ndahl, Christiania, 1881; the five-term relation is in the note on the dilogarithm.
\bibitem{rogers} L.~J.~Rogers, On function sum theorems connected with the series
$\sum x^n/n^2$, \emph{Proc.\ London Math.\ Soc.} \textbf{4} (1907) 169--189.
\bibitem{lewin} L.~Lewin, \emph{Polylogarithms and Associated Functions}, North-Holland,
1981.
\bibitem{zagier} D.~Zagier, The dilogarithm function, in \emph{Frontiers in Number Theory,
Physics, and Geometry II}, Springer, 2007, 3--65.
\bibitem{wojtkowiak} Z.~Wojtkowiak, The basic structure of polylogarithmic functional
equations, in \emph{Structural Properties of Polylogarithms} (L.~Lewin, ed.), Math.\
Surveys Monogr.\ \textbf{37}, AMS, 1991, 205--231.
\bibitem{dejeu} R.~de~Jeu, Towards regulator formulae for the $K$-theory of curves and
number fields, \emph{Compositio Math.} \textbf{124} (2000) 137--194.
\bibitem{bloch} S.~Bloch, \emph{Higher Regulators, Algebraic $K$-theory, and Zeta Functions
of Elliptic Curves}, CRM Monogr.\ Ser.\ \textbf{11}, AMS, 2000 (Irvine lectures, 1978).
\bibitem{suslin} A.~A.~Suslin, $K_3$ of a field, and the Bloch group, \emph{Proc.\ Steklov
Inst.\ Math.} \textbf{183} (1991) 217--239.
\bibitem{neumannzagier} W.~D.~Neumann, D.~Zagier, Volumes of hyperbolic three-manifolds,
\emph{Topology} \textbf{24} (1985) 307--332.
\bibitem{goncharov} A.~B.~Goncharov, Geometry of configurations, polylogarithms, and
motivic cohomology, \emph{Adv.\ Math.} \textbf{114} (1995) 197--318.
\bibitem{nakanishi} T.~Nakanishi, Dilogarithm identities for conformal field theories and
cluster algebras, in \emph{Symmetries, Integrable Systems and Representations}, Springer
Proc.\ Math.\ Stat.\ \textbf{40} (2013) 407--444.
\bibitem{faddeevkashaev} L.~D.~Faddeev, R.~M.~Kashaev, Quantum dilogarithm as a kernel of
the quantum modular double, \emph{Modern Phys.\ Lett.\ A} \textbf{9} (1994) 427--434.
\bibitem{kirillov} A.~N.~Kirillov, Dilogarithm identities, \emph{Prog.\ Theor.\ Phys.\
Suppl.} \textbf{118} (1995) 61--142.
\bibitem{zamolodchikov} Al.~B.~Zamolodchikov, Thermodynamic Bethe ansatz in relativistic
models, \emph{Nucl.\ Phys.\ B} \textbf{342} (1990) 695--720.
\bibitem{nahm} W.~Nahm, Conformal field theory and torsion elements of the Bloch group, in
\emph{Frontiers in Number Theory, Physics, and Geometry II}, Springer, 2007, 67--132.
\bibitem{clock} C.~Mar\'in, \emph{The Clock That Never Ticks: a machine-checked path from
the dilogarithm to quantum speed limits in Lean~4}, companion paper, same repository, 2026.
\bibitem{mathlib} The mathlib Community, The Lean mathematical library, in \emph{CPP 2020},
ACM, 2020, 367--381.
\end{thebibliography}

\end{document}
