\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{hyperref}
\usepackage{enumitem}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\dart}{\mathrm{Dart}}
\newcommand{\detm}{\det}
\DeclareMathOperator{\charpoly}{charpoly}

\title{The Ihara Side: A Machine-Checked Proof of\\ Bass's Determinant Formula in Lean 4}
\author{Carles Mar\'in\thanks{Independent researcher, \texttt{karlesmarin@gmail.com}. The
formalization was carried out with Claude (Anthropic) as an AI research instrument under a
build-as-oracle loop: the assistant proposed definitions and proofs, the Lean kernel verified
or rejected each, and all design decisions, mathematics and claims are the author's. The kernel
has the last word.}}
\date{2026}

\begin{document}
\maketitle

\begin{abstract}
We give a machine-checked proof, in Lean~4 / Mathlib, of \emph{Bass's determinant formula}
for the Ihara zeta function of a finite graph:
\[
  (1-u^2)^{\lvert V\rvert}\,\det\!\big(I - u\,B\big)
   \;=\;(1-u^2)^{\lvert E\rvert}\,\det\!\big(I - u\,A + u^2(D-I)\big),
\]
where $B$ is Hashimoto's non-backtracking (edge-adjacency) operator on the $2\lvert E\rvert$
oriented edges of the graph, $A$ is the adjacency matrix and $D$ the degree matrix. Equivalently
$\det(I-uB)=(1-u^2)^{\lvert E\rvert-\lvert V\rvert}\det(I-uA+(D-I)u^2)$, the reciprocal of the
Ihara zeta function. To our knowledge this is the first formalization of Bass's formula in any
proof assistant. All headline theorems are \texttt{sorry}-free: \texttt{\#print axioms} reports
only \texttt{propext}, \texttt{Classical.choice}, \texttt{Quot.sound}. The proof is over a field,
the standard setting for the identity. This is the \emph{Ihara / cycle} side complementing a prior
Lean formalization of the matching polynomial (the tree / Plancherel side).
\end{abstract}

\section{Introduction}

The Ihara zeta function of a finite graph $G$, introduced by Ihara in the 1960s and recast for
graphs by Serre, Sunada, Hashimoto and Bass, is
\[
  \zeta_G(u)^{-1} \;=\; \det\!\big(I_{2\lvert E\rvert} - u\,B\big),
\]
where $B$ is the \emph{non-backtracking} (Hashimoto) operator on the directed edges: $B_{d,e}=1$
exactly when the head of $d$ is the tail of $e$ and $e$ is not the reversal of $d$. Bass's
theorem reduces this $2\lvert E\rvert$-dimensional determinant to an $\lvert V\rvert$-dimensional
one built from the adjacency and degree matrices. The non-backtracking spectrum governs, among
other things, percolation and community-detection thresholds on networks; Bass's identity is the
bridge between the combinatorial (edge) and spectral (vertex) descriptions.

We formalize the finite, division-free form of the identity. We deliberately do \emph{not}
formalize the analytic zeta function (the Euler product over prime cycles): the determinant
identity is the content with an elementary proof, and it is what no proof assistant had recorded.

\paragraph{Honesty note.} This is a formalization of classical mathematics (Bass, 1992); it proves
no new theorem and claims none. The ``first formalization'' claim rests on a search of the
Lean/Mathlib, Isabelle/AFP and Coq ecosystems, not on an exhaustive byte-level audit.

\section{The formalization}

Let $G$ be a finite simple graph with vertex type $V$ (a \texttt{Fintype} with decidable
adjacency). Mathlib's \texttt{SimpleGraph.Dart} is the type of oriented edges (darts); reversal
$d\mapsto d^{\mathrm{symm}}$ is a fixed-point-free involution on the $2\lvert E\rvert$ darts. Over
a commutative ring $R$ we define, as \texttt{Matrix} valued objects:
\begin{itemize}[nosep]
  \item the degree matrix $D=\texttt{degMatrix}=\operatorname{diagonal}(v\mapsto \deg v)$;
  \item the reversal permutation matrix $J=\texttt{reversal}$, $J_{d,e}=[\,e=d^{\mathrm{symm}}\,]$;
  \item the Hashimoto operator $B=\texttt{hashimoto}$,
        $B_{d,e}=[\,d.\mathrm{snd}=e.\mathrm{fst}\ \wedge\ e\neq d^{\mathrm{symm}}\,]$;
  \item incidence matrices $S=\texttt{startInc}$, $T=\texttt{termInc}$ ($\lvert V\rvert\times2\lvert E\rvert$),
        $S_{v,d}=[\,d.\mathrm{fst}=v\,]$, $T_{v,d}=[\,d.\mathrm{snd}=v\,]$;
\end{itemize}
and $A=$ Mathlib's \texttt{adjMatrix}. The statement proved is
\begin{theorem}[\texttt{bass\_determinant}]
Over a field $R$, for $u\in R$,
\[
  (1-u^2)^{\lvert V\rvert}\detm(1 - u\cdot B)
   = (1-u^2)^{\lvert E\rvert}\detm\!\big(1 - u\cdot A + u^2\cdot(D-1)\big).
\]
\end{theorem}

\section{Outline of the proof}

\paragraph{Incidence layer.} Five identities, each a finite sum over darts (all numerically
pre-verified before formalization):
\[
  S\,T^{\mathsf T}=A,\quad S\,S^{\mathsf T}=D=T\,T^{\mathsf T},\quad
  B = T^{\mathsf T}S - J,\quad S\,J = T.
\]
The dart-counting lemma $\#\{d : d.\mathrm{fst}=v,\ d.\mathrm{snd}=w\}=[\,v\sim w\,]$ realizes the
adjacency matrix.

\paragraph{$\det(I+uJ)=(1-u^2)^{\lvert E\rvert}$.} Fixing a linear order on $V$, a dart is
\emph{positive} when its tail precedes its head; the reversal of a non-positive dart is positive.
This yields an equivalence
\[
  \texttt{dartEquiv}:\ \dart \;\simeq\; \mathrm{Bool}\times\{\text{positive darts}\},
\]
under which $J$ reindexes to a block-diagonal of the $2\times2$ swap, so $I+uJ$ becomes a
block-diagonal of $\left[\begin{smallmatrix}1&u\\u&1\end{smallmatrix}\right]$ (one block per edge,
since $\lvert\{\text{positive darts}\}\rvert=\lvert E\rvert$ via the handshake lemma). Hence
$\det(I+uJ)=(1-u^2)^{\lvert E\rvert}$ by \texttt{Matrix.det\_blockDiagonal}.

\paragraph{Sylvester step ($1-u^2\neq0$).} Since $B=T^{\mathsf T}S-J$, we have
$I-uB=(I+uJ)-u\,T^{\mathsf T}S$. Over a field with $1-u^2\neq0$, $I+uJ$ is invertible with
$(I+uJ)^{-1}=(1-u^2)^{-1}(I-uJ)$ (from $(I+uJ)(I-uJ)=(1-u^2)I$). Factoring it out and applying the
Weinstein--Aronszajn identity (\texttt{Matrix.det\_one\_sub\_mul\_comm}) to swap $S$ past
$T^{\mathsf T}$ gives
\[
  \det(I-uB)=(1-u^2)^{\lvert E\rvert}\,\det\!\big(I - u\,S(I+uJ)^{-1}T^{\mathsf T}\big),
\]
and $S(I-uJ)T^{\mathsf T}=A-uD$ collapses the right-hand determinant to
$(1-u^2)^{-\lvert V\rvert}\det(I-uA+u^2(D-I))$. Clearing denominators yields the theorem.

\paragraph{Degenerate value $u^2=1$.} Then $1-u^2=0$; the claim reduces to the no-edge and
empty-graph cases, where the matrices are scalar multiples of the identity or live over an empty
index, and both sides vanish or are trivially equal.

\begin{remark}
The field hypothesis is the usual setting for Bass's formula. The inverse-free manipulations alone
yield the identity multiplied by $(1-u^2)^{\lvert E\rvert}$; recovering the tight form requires
cancelling that factor, which a field permits. Full commutative-ring generality follows by a
universal-coefficient transfer (prove over $\mathbb{Z}[u]$, a domain, then base-change), not pursued
here.
\end{remark}

\section{Relation to the matching polynomial}

This development is the \emph{Ihara / $\pi_1$ side} of a graph trace formula whose \emph{tree /
Plancherel side} is the matching polynomial, formalized in the companion papers \emph{Random Signs
into Matchings} and \emph{Unfolding a Graph into a Tree}. With both endpoints now machine-checked,
a future work can formalize the trace formula itself: $\operatorname{tr}(A^k)=p_k+N_k$, the
matching (tree-like) term plus the non-backtracking (cycle) term.

\section*{Availability}

The Lean source is at \url{https://github.com/karlesmarin/godsil-gutman-lean}
(\texttt{Ihara/Bass.lean}), pinned to Lean \texttt{v4.30.0-rc2} and a fixed Mathlib revision.
Verify the axiom footprint with
\[
  \texttt{\#print axioms SimpleGraph.bass\_determinant}
   \;\rightsquigarrow\;\texttt{[propext, Classical.choice, Quot.sound]}.
\]

\end{document}
